---
title: "ACME Detection Figures for 2023-2024 OSM Fiscal Report"
author: "Ma"
date: "2025-03-09"
output: 
  html_document:
    theme: default
    toc: yes
    toc_float: yes
---


The first two chunks of this r markdown file after the r setup allow for plot zooming, but it also means that the html file must be opened in a browser to view the document properly. When it knits in RStudio the preview will appear empty but the html when opened in a browser will have all the info and you can click on each plot to Zoom in on it. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Before you begin

## Notes

A few notes about this script.

If you are running this with the 2023-2024 data make sure you download the whole (OSM_2023-2024 GitHub repository)[https://github.com/ACMElabUvic/OSM_2023-2024] from the ACMElabUvic GitHub. This will ensure you have all the files, data, and proper folder structure you will need to run this code and associated analyses.

Also make sure you open RStudio through the R project (OSM_2023-2024.Rproj) this will automatically set your working directory to the correct place (wherever you saved the repository) and ensure you don't have to change the file paths for some of the data.

Lastly, if you are looking to adapt this code for a future year of data, you will want to ensure you have run the 1_ACME_camera_script_2025-02-07.R or .Rmd with your data as there is much data formatting, cleaning, and restructuring that has to be done before this code will work. *Helpful note: The files are numbered in the order they are used for this analysis*.

If you have question please email the most recent author, currently   

Aidan Brushett
M.Sc. Student
University of Victoria    
School of Environmental Studies     
Email: [aidanbrushett@uvic.ca](aidanbrushett@uvic.ca)     

(*update/add authors as needed*)


## R and RStudio

Before starting you should ensure you have the latest version of R and RStudio downloaded. This code was generated under R version 4.2.3 and with RStudio version 2024.04.2+764.    

You can download R and RStudio [HERE](https://posit.co/download/rstudio-desktop/)   


## R markdown

This script is written in R markdown and thus uses a mix of coding markup languages and R. If you are planning to run this script with new data or make any modifications you will want to be familiar with some basics of R markdown.

Below is an R markdown cheatsheet to help you get started,    
[R markdown cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)    


## Install and load packages

If you don't already have the following packages installed, use the code below to install them. *NOTE this will not run automatically as eval=FALSE is included in the chunk setup (i.e. I don't want it to run every time I run this code since I have the packages installed)

```{r install packages, eval=FALSE}
#install.packages('tidyverse')
#nstall.packages('PerformanceAnalytics')
#install.packages('Hmisc')
```

Load your needed packages: 

I will also make a sneaky function that I love. 
```{r}
library(tidyverse)

`%nin%` <- Negate(`%in%`)
```

# 1. Import data

Let's read in the independent detections data for all seasons. We'll also clean up the array names by removing the leading zeros (I believe this was the standard we wanted moving forward). The `str_replace()` solution is a bit of a brute fix but it works just fine. 

```{r}
indet <-    map_dfr(list.files(path = "./data/processed/", pattern = "OSM_ind_det", full.names = T),
                    ~ read_csv(.x, 
                               col_types = cols(array = col_factor(),
                                           site = col_factor(),
                                           datetime = col_datetime(),
                                           event_id = col_character()
                                           )
                               )
                    ) %>%
  
  select(-event_id, -timediff) %>%
  
  mutate(site = str_replace_all(site, c("LU09" = "LU9",
                                        "LU01" = "LU1")) %>% 
           as.factor(.),
         
         array = str_replace_all(array, c("LU09" = "LU9",
                                          "LU01" = "LU1")) %>% 
           as.factor(.)
         )

levels(indet$array)
levels(indet$site)
```

We'll also reformat this into presence absence for each species and site. This is very coarse and does not reflect data availability, sampling effort, etc. but since this is an exploratory figure I won't loose too much sleep over it.  The code to make figures was adapted from my thesis code and might be slightly redundant but it's still relatively legible and works for the purposes of the report. 

```{r}
presence_temp <- indet %>%
  
  group_by(array, site, species) %>%
  
  summarize(presence = 1)


presence <- crossing(species = unique(indet$species), site = unique(indet$site)) %>%
  
  full_join(presence_temp[1:2], by = 'site') %>%
  
  left_join(presence_temp[2:4], by = c('species', 'site')) %>%
  
  mutate(presence = ifelse(is.na(presence), 0, 1)) %>%
  
  distinct()
```

Finally, we'll load up the LU classifications from the BADR report. 

```{r}
lu_classifications <- read_csv("./data/raw/landscape_unit_land_use_classifications.csv")
```


# 2. Make some figures

## 2.0. Identify focal species

We'll identify some focal species (not used for now) and also bin some of the similar species categories. 

```{r}
arrays <- unique(indet$array) 

c('LU9', 'LU22', 'LU14', 'LU16')

# We will not use this for now
focal_species <- c('Red squirrel',
                   'Snowshoe hare', 
                   'White-tailed deer',
                   'Black bear',
                   'Coyote',
                   'Moose',
                   'Lynx', 
                   'Unknown',
                   'Marten',
                   'Wolverine',
                   'Grey wolf',
                   'Caribou',
                   'Red fox', 
                   'Fisher',
                   'Spruce grouse',
                   'Ruffed grouse', 
                   'Grey jay',
                   'Long-tailed weasel',
                   'Ruffed grouse',
                   'Short-tailed weasel',
                   'Otter',
                   'Canada goose',
                   'Beaver',
                   'Human (various)',
                   'Porcupine',
                   'Owl',
                   'Cougar'
                   )

human_species <- c('ATVer', 
                   'Human',
                   'Hunter',
                   'Hiker',
                   'Domestic dog',
                   'Snowmobiler'
                   )

unknown_species <- c('Unknown',
                     'Unknown canid', 
                     'Unknown ungulate',
                     'Unknown deer',
                     'Unknown mustelid',
                     'Unknown bear')
```


## 2.1. Independent detections and naive occupancy per array

For now we'll only plot the 2023-24 arrays. This will be LU9, LU9, LU14, LU16, and LU22. We read in everything above-- in case this changes in future
```{r}
# we can just work with these arrays for now
for(LU in arrays){
  
  fig_indet <- indet %>%
    
    # Group together a few similar species
    mutate(species = case_when(
      species %in% human_species ~ "Human (various)",
      species %in% unknown_species ~ "Unknown (various)",
      TRUE ~ species)) %>%
    
    # Remove the species/arrays we don't want in the figure
    filter(array == LU,
           !species %in% c("Staff", 
                           "Other", 
                           "Other birds")) %>%
    
    group_by(species) %>%
    
    # How many rows (detections) per species?
    summarize(total = n()) %>%
    
    # Ensure species is ordered by total detections
    mutate(species = fct_reorder(species, as.numeric(total))) %>%
    
    ggplot(., aes(x = total, y = species)) +
    
      geom_bar(stat = "identity", fill = "gray30") +
    
      # Add labels to each bar
      geom_text(aes(label = total), hjust = -0.3, size = 2) +  # Adjust hjust for positioning
  
    
      labs(x = "Number of independent (30 min) detections",
           y = "",
           title = paste0(LU, 
                          ' (', 
                          subset(lu_classifications, array == LU)$classification,
                          ')')
           ) + 
    
      scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +  # Removes whitespace on the left

  theme_classic()

  fig_indet
  
  # Save the base figure
  ggsave(paste0("./figures/", LU, "_independent_detections.png"), width = 8, height = 5, dpi = 500) 

  
  # Now naive occupancy
  fig_naiveocc <- presence %>%
  
    # Just the array for the figure
    filter(array == LU) %>%
  
    # Group together a few similar species
    mutate(species = case_when(
      species %in% human_species ~ "Human (various)",
      species %in% unknown_species ~ "Unknown (various)",
      TRUE ~ species)) %>%
  
    group_by(array, species) %>%
  
    # Within an array, how many sites? How many present? Naive = proportion present. 
    summarize(n_sites = n(), 
              n_present = sum(presence),
              naive_occ = n_present / n_sites,
              .groups = 'drop') %>%  
  
    # Remove the species we don't want in the figure
    filter(!species %in% c("Staff", 
                           "Other", 
                           "Other birds"),
           naive_occ > 0) %>%
    
    # Ensure species is ordered by total detections
    mutate(species = fct_reorder(species, as.numeric(naive_occ))) %>%
  
    ggplot(., aes(x = naive_occ, y = species)) +
    
      geom_bar(stat = "identity", fill = "gray30") +
    
      # Add labels to each bar
      geom_text(aes(label = round(naive_occ, 2)), hjust = -0.3, size = 2) +  # Adjust hjust for positioning
  
      labs(x = "Naive occupancy",
           y = "",
           title = "   ") +
    
      scale_x_continuous(expand = expansion(mult = c(0, 0.05))) + # Removes whitespace on the left
    
    theme_classic()

fig_naiveocc

# Combine the figures into one
ggpubr::ggarrange(fig_indet, 
                  fig_naiveocc,
                  ncol = 1)

# Save the combined figure
ggsave(paste0("./figures/", LU, "_combined_detections_occupancy.png"), width = 8, height = 10, dpi = 500) 
}
# end for loop
```


This is the exact same code as above (for the naive occupancy) but with a title added. I'm super tired so just brute forcing this code. 

```{r}
# we can just work with these arrays for now
for(LU in arrays){
  
  # Now naive occupancy
  fig_naiveocc <- presence %>%
  
    # Just the array for the figure
    filter(array == LU) %>%
  
    # Group together a few similar species
    mutate(species = case_when(
      species %in% human_species ~ "Human (various)",
      species %in% unknown_species ~ "Unknown (various)",
      TRUE ~ species)) %>%
  
    group_by(array, species) %>%
  
    # Within an array, how many sites? How many present? Naive = proportion present. 
    summarize(n_sites = n(), 
              n_present = sum(presence),
              naive_occ = n_present / n_sites,
              .groups = 'drop') %>%  
  
    # Remove the species we don't want in the figure
    filter(!species %in% c("Staff", 
                           "Other", 
                           "Other birds"),
           naive_occ > 0) %>%
    
    # Ensure species is ordered by total detections
    mutate(species = fct_reorder(species, as.numeric(naive_occ))) %>%
  
    ggplot(., aes(x = naive_occ, y = species)) +
    
      geom_bar(stat = "identity", fill = "gray30") +
    
      # Add labels to each bar
      geom_text(aes(label = round(naive_occ, 2)), hjust = -0.3, size = 2) +  # Adjust hjust for positioning
  
      labs(x = "Naive occupancy",
           y = "",
           title = paste0(LU, 
                          ' (', 
                          subset(lu_classifications, array == LU)$classification,
                          ')')
           ) + 
    
      scale_x_continuous(expand = expansion(mult = c(0, 0.05))) + # Removes whitespace on the left
    
    theme_classic()

fig_naiveocc

# Save the base figure
ggsave(paste0("./figures/", LU, "_naive_occupancy.png"), width = 8, height = 5, dpi = 500) 

} # end for loop
```
# End of script
